Programming Language Survey Paper on Haskell

### 2013-04-22 projects/term_paper/paper.tex
More edits.

\usepackage{appendix}
\usepackage{minted}

@@

Haskell is named after Haskell Curry, a logician whose work greatly contributed to functional programming. The logo of Haskell, lambda,
symbolize lambda calculus as its foundation. Today, Haskell is widely used in many different aspects 
of computing.

@@

@@

Haskell is most commonly used for the development of substantial software systems. The language is designed to increase
productivity from programmers in fewer lines of code. In addition, having no side-effects leads to fewer errors.
These features are very important in the development of large software systems, and Haskell is an ideal choice.

@@

\subsection{Data Abstractions}
Haskell supports a very rich, strong, static type system for programmers.  As Haskell is
a compiled language, with the Glasgow Haskell Compiler (GHC) being the most common implementation,
most of the type checking is performed and enforced at compile time.

@@

Haskell also allows the definition Strings, which are represented by surrounding text with double quotes. Strings, however, 
are not their own data type. Although string literals can be initialized using the double quote syntax, the compiler converts the 
string into a list of characters at compile time. The use of the String notation is simply syntatic sugar for programmers. 
It also helps increase the readability of code. In the following example, \texttt{str1} shows how the data for a string is actually stored: as a list of characters. However, Haskell allows the definition of a string using double quotes, as seen in \texttt{str2}, 
which is much more understandable.

@@

In addition to these basic data types, Haskell also offers type variables. Type variables are variables whose type has 
not been explicitly declared; rather, they can be any type.  By defining a function without declaring data types, the 
variables in the function can be used as any type thereby enabling polymorphic behaviors. In the following example, the 
function can handle a list of any type because its uses a type variable.

@@

\subsubsection{Type Checking}
Type checking in Haskell is performed statically at compile time. Haskell uses type-declarations for functions, in 
which the types of parameters and return values are explicitly defined. Any use of a function in which parameters 
do not meet the defined function type is flagged by the compiler, and compilation will be aborted. In addition, 
Haskell uses type inferencing by context. The type system in Haskell deduces obvious types and makes them concrete. In the 
following example, although the type of myNum is unknown at compile time, the compiler infers that the Int type 
is desired because myNum is added to an Int.

\begin{minted}{haskell}
let myStr = "2" 
let myNum2 = (read myStr) + 2
\end{minted}

@@

Haskell uses typeclasses to categorize the type system. Typeclasses define the behavior of a general type. 
A type must support the definition of its typeclass, but may also be more specific. For example, all of the 
operators in Haskell are categorized under the Eq typeclass. Other examples of typeclasses in Haskell include Ord, 
Show, Read, Enum, Num, Integral, and Floating. Typeclasses can be used to allow multiple types to be passed to a 
function. In the following example, the function declares its typeclass as Integral. Therefore, the function will 
allow accept Int or Integer types as parameters.

@@

\subsection{Control Abstractions}
\subsubsection{Expressions}

@@

The second type of binding available in Haskell is done using \texttt{let} bindings. \texttt{let} bindings allow the declaration of 
variables that are expressions themselves and are locally bound.  \texttt{let} bindings are often used for pattern matching. 
The syntax of \texttt{let} bindings involves declaring the bindings inside a \texttt{let} declaration, followed by defining an 
expression inside an \texttt{in} declaration. The \texttt{let} and \texttt{in} statements must be aligned in the same column.

\begin{minted}{haskell}
myExp :: (Num a) => a -> a
myExp =
    let e = 2.7182
    in  e ** x
\end{minted}

The difference between \texttt{let} and \texttt{where} bindings is that \texttt{let} bindings are expressions themselves, whereas 
\texttt{where} bindings are just syntactic constructs. Additionally, unlike \texttt{where} bindings, \texttt{let} bindings cannot 
span across guards.

@@

If-else statements are not used in Haskell as often as they are in imperative languages because Haskell offers a feature 
known as guards, which are fundamentally equivalent to if-else statements except they are much more easily understood, 
work with multiple conditions, and blend well with pattern matching. Guards are created using pipes in the 
expression of a function and consist of a conditional, which must evaluate to true or false, and an expression. 
An equals sign separates the conditional from the expression. Arguments are checked against the guards in 
sequential order. If the condition is true, then the expression for that guard is evaluated and returned. 
If the condition is false, then the next guard condition is evaluated, and this pattern continues until the 
\texttt{otherwise} guard is reached, which catches all input that did not match any guard previous conditions. The 
following function has three guards, each following a pipe. A notable aspect of this function is that if the 
input is value is greater than 100,000 it will cause the first guard condition to be true and will evaluate the 
first guard expression even though the input would also be true for the second guard. This is because the guards are 
checked in the order which they are defined, so the second guard is never checked in this case because the first guard is true and the return value is 
determined from it. The \texttt{otherwise} guard is placed at the bottom to catch any value that does not cause the first two guards to be true.

\begin{minted}{haskell}
let balance = 200
assessWorth :: Int -> String
assessWorth =
    | balance > 100000 = "You're rich!"
    | balance > 100 = "You're rich by student standards!"
    | otherwise = "Need a loan, friend?"
\end{minted}

@@

\subsubsection{Functions}

@@

For pure functions, the Haskell typeclass system was broadened to allow the introduction
of the \texttt{Maybe} algebraic type.  This type extension provides a function a way to
indicate that it may or may not return a value of the specified type in its declaration
(i.e., it may fail to return any value).  If a function implementing this typeclass
fails this way, the return value will be \texttt{Nothing}.  If it does return a value, say \texttt{x},
that value will be prefaced by the keyword \texttt{Just}.  A pure function illustrating this concept
is given below.
\begin{minted}{haskell}
doubleHead :: [Int] -> Maybe Int
doubleHead [] = Nothing
doubleHead xs = Just ((head xs) * 2)

>>>>

\usepackage{appendix}
\usepackage{minted}
\usepackage{url}

@@

Haskell is named after Haskell Curry, a logician whose work greatly contributed to functional programming. The logo of Haskell, lambda,
symbolizes that lambda calculus is the foundation from which Haskell was based. Today, Haskell is widely used in many different aspects 
of computing.

@@

Beside function independence, Haskell also follows the lazy function evaluation paradigm which characterizes some other functional 
languages. Because of lazy evalution, functions in Haskell are not evaluated until their return value is needed by some other function.
This leads to some interesting constructs that cannot be implemented in non-lazy languages, such as infinite lists.

@@

Haskell is most commonly used for the development of substantial software systems. The language is designed to increase
productivity from programmers in fewer lines of code. In addition, fewer errors are often produced by developers due to the lack of
side-effects in functions.  These features are very important in the development of large software systems, and Haskell is an ideal choice.

@@

\subsection{Data Abstractions}
Haskell supports a very rich, strong, static type system for programmers.  As Haskell is
a compiled language --- with the Glasgow Haskell Compiler (GHC) being the most common implementation ---
most of the type checking is performed and enforced at compile time.

@@

Haskell also allows definition Strings, which are represented by surrounding text with double quotes. Strings, however, 
are not their own data type. Although string literals can be initialized using the double quote syntax, the compiler converts the 
string into a list of characters at compile time. The use of the String notation is simply syntatic sugar for programmers. 
It also helps increase the readability of code. In the following example, \texttt{str1} shows how the data for a string is actually stored: as a list of characters. However, Haskell allows the definition of a string using double quotes, as seen in \texttt{str2}, 
which is much more understandable.

@@

In addition to these basic data types, Haskell also offers type variables. Type variables are variables whose type has 
not been explicitly declared; rather, they can be any type.  Type variables are most often employed to define the parameter
and return value types of functions. By defining a function without declaring data types, the 
variables in the function can be used as any type thereby enabling polymorphic behaviors. In the following example, the 
function can handle a list of any type because its uses a type variable.

@@

\subsubsection{Type Checking}
Type checking in Haskell is performed statically at compile time. Haskell uses type-declarations for functions, in 
which the types of parameters and return values are explicitly defined. Any use of a function in which parameters 
do not meet the defined function type is flagged by the compiler, and compilation will be aborted. In addition, 
Haskell supports type inferencing by context. The type system in Haskell deduces obvious types and makes them concrete. In the 
following example, the \texttt{read} function is used to convert \texttt{myStr} to another data type.
Although the type of \texttt{myNum} is unknown at compile time, the compiler infers that the Int type 
is desired because \texttt{myStr} is added to an Int type.

\begin{minted}{haskell}
let myStr = "2" 
let myNum = (read myStr) + 2
\end{minted}

@@

To avoid ambiguity in type conversion, the following syntax can be employed to indicated to the compiler what type
is desired from the \texttt{read} function.

\begin{minted}{haskell}
let myStr = "2"
let myNum = (read myStr) :: Float
\end{minted}

Haskell uses typeclasses to categorize the type system. Typeclasses define the behavior of a general type. 
A type must support the definition of its typeclass, but may also be more specific. For example, all of the 
operators in Haskell are categorized under the Eq typeclass. Other examples of typeclasses in Haskell include Ord, 
Show, Read, Enum, Num, Integral, and Floating. Typeclasses can be used to allow multiple types to be passed to a 
function. In the following example, the function declares its typeclass as Integral. Therefore, valid functions invocations consist
of Int or Integer type parameters.

@@

\subsection{Control Abstractions}

Haskell supports several constructs for controlling program execution.  These constructs can be further divided into:
expressions, functions, and operators, selection and iteration mechanisms, modules, and execption handling.
In order to further understand the functionality of these constructs, scoping and parameter passing techniques are discussed.

\subsubsection{Expressions and Functions}

@@

The second type of binding available in Haskell is done using \texttt{let} bindings. \texttt{let} bindings allow the declaration of 
variables that are expressions themselves and are locally bound.  \texttt{let} bindings are often used for pattern matching. 
The syntax of \texttt{let} bindings involves declaring the bindings inside a \texttt{let} declaration, followed by defining an 
expression inside an \texttt{in} declaration. The expression in the \texttt{let} and \texttt{in} parts must be aligned in the same column.

\begin{minted}{haskell}
myExp :: (Num a) => a -> a
myExp x =
    let e = 2.7182
    in  e ** x
\end{minted}

The difference between \texttt{let} and \texttt{where} bindings is that \texttt{let} bindings are expressions themselves, whereas 
\texttt{where} bindings are just syntactic constructs attached to some expresion. 
Additionally, unlike \texttt{where} bindings, \texttt{let} bindings cannot span across guards.

@@

If-else statements are not used in Haskell as often as they are in imperative languages because Haskell offers another feature 
known as guards, which are fundamentally equivalent to if-else statements except they are much more easily understood, 
work with multiple conditions, and blend well with pattern matching. Guards are created using pipes in the 
expression of a function and consist of a conditional, which must evaluate to true or false, and an expression. 
An equals sign separates the conditional from the expression. Arguments are checked against the guards in 
sequential order. If the condition is true, then the expression for that guard is evaluated and returned. 
If the condition is false, then the next guard condition is evaluated, and this pattern continues until the 
\texttt{otherwise} guard is reached, which catches all input that did not match any guard previous conditions. The 
following function has three guards, each following a pipe. A notable aspect of this function is that if the 
input is value is greater than 100,000 it will cause the first guard condition to be true and will evaluate the 
first guard expression even though the input would also be true for the second guard. This is because the guards are 
checked in the order which they are defined, so the second guard is never checked in this case because the first guard is true and 
the return value is  determined from it. The \texttt{otherwise} guard is placed at the bottom to catch any value that does not 
cause the first two guards to be true.

\newpage
\begin{minted}{haskell}
let balance = 200
assessWorth :: Int -> String
assessWorth =
    | balance > 100000 = "You're rich!"
    | balance > 100 = "You're rich by student standards!"
    | otherwise = "Need a loan, friend?"
\end{minted}

@@

\subsubsection{Functions and Parameter Passing}

@@

For pure functions, the Haskell typeclass system was broadened to allow the introduction
of the \texttt{Maybe} algebraic type.  This type extension provides a function a way to
indicate that it may or may not return a value of the specified type in its declaration
(i.e., it may fail to return any value).  If a function implementing this typeclass
fails this way, the return value will be \texttt{Nothing}.  If it does return a value, say \texttt{x},
that value will be prefaced by the keyword \texttt{Just}.  A pure function illustrating this concept
is given below.
\newpage
\begin{minted}{haskell}
doubleHead :: [Int] -> Maybe Int
doubleHead [] = Nothing
doubleHead xs = Just ((head xs) * 2)
